Redshift evolution of the SN stretch distribution

### 2019-11-27 Nicolas.tex
Finished zmax inputs

\documentclass{aa}
\usepackage{preambule}
\usepackage{graphicx}
\usepackage[varg]{txfonts}

@@

\author{N. Nicolas\inst{1}
    \and M. Rigault\inst{2}
}

@@

\begin{figure*}[htbp!]
    \centering
    \subfigure[Method]{\label{fig:zdet}
        \includegraphics[width=8.5cm]{Article_figures/zmax_determination.pdf}}
    \subfigure[Result]{\label{fig:surveys_cuts}
    \includegraphics[width=8.5cm]{Article_figures/surveys_cuts_x.pdf}}
    \caption{(a) $z\St{max}$ determination from the interpolated poissonian
    evolutions; (b) SDSS, SNLS and PS1 samples cut at $z\St{max}$ as explained in
    section \ref{sec:sam}. The data we use is in plain markers.}
\end{figure*}

Type Ia supernovae (SNe Ia) are now well-known for their capacity to determine
cosmological parameters: their study led to the discovery of the accelerated
expansion of the Universe (Riess 98, Perlmutter 99) through the name of "dark
energy", and they have been used continuously for better measurments since then
(Betoule 2014). They are acquired through their lightcurves, giving the
evolution of their luminosity from the time of explosion, in different
wavelength. 3 parameters are used to describe those: an amplitude, a width
(named "stretch") and a color (magnitude difference in the B and V bands).

The simple use of those is not enough for our aim, as SNe Ia have an intrinsec
dispersion of their luminosity of $\approx 0.4~mag$ that gives a huge
uncertainty on the determination of their distance modulus. Henceforth they are
standardized using the "brigther-slower" and "brighter-bluer" relations
(Philipps 93, Riess 96, Tripp 98) in the SALT2 algorithm (Guy 2007, 2010) that
fits the distance modulus which is expressed as
\begin{equation}
    \m = m_b + \a x_1 - \b c - M
\end{equation}
with $m_B$ the logarithm of their flux, $x_1$ their stretch, $c$ their color and
$M$ their intrinsec magnitude. This relation lowered their magnitude dispersion
to $\approx 0.15~mag$, allowing for previously mentioned accelerated
expansion to be discovered.

This Tripp estimator lies on the idea that this standardization doesn't change
with the redshift. However, Rigault 2015 showed that SNe Ia depend on their
environments, and these environments' properties evolve with the redshift. In
this Letter, we try to determine whether a stretch evolution with the redshift
allows for a better description of the collected data.

@@

We use data from 5 different surveys: Hubble Space Telescope (HST, REF),
Supernova Nearby Factory (SNf, REF), Supernova Legacy Survey (SNLS, Aster 2006,
Betoule 2014), Sloan DSS (SDSS, REF) and Panstarr-1 (PS1, REF). The last 3 had
selection effects that come from the "brighter-slower" and "brighter-bluer"
relations: the first SNe Ia that are missed by the instruments are the redder
and faster ones. That implies that taking all of these surveys' data would not
represent the stretch distribution that Nature could give us.

As such, we determined the maximum redshift at which we would cut these samples
to get complete ones. We used a statistical approach in lack of precise data
concerning the instruments' capacity to acquire fluxes. We proceeded by making
histograms of the surveys, and defining a function of the expected SNe Ia rate:
\begin{equation}
    N\St{SNe Ia} = a \times V\z
\end{equation}
with $a$ the density of SNe Ia and $V\z$ the volume of Universe in each bin.
Before selection effects, the histograms are supposed to grow proportionnaly to
$V\z$ assuming an homogeneous repartition of SNe Ia. When we start to miss some
supernovae, this growth is not followed anymore.

To determine this limit, we first took the observed counts in each bin of the
histograms. We then fitted $a$ on these counts for different numbers of bins
used to fit: the quality of the fit diminishes when we use bins that include
selection effects. To quantize this, we computed the expected counts from the
fitted $a$ in the same bins we used for the observed counts. That allowed us to
compute the poissonian probability in each bin, given by
\begin{equation}
    p(k) = \frac{\l^k}{k!}e^{-\l}
\end{equation}
with $k$ the expected counts, $\l$ the observed counts. This work is done 100
times with a random number of total bins, $a$ being fitted 10 times on a
different and random number of bins used to fit each time. We interpolate these
1000 poissonian evolutions to get one final result that is as close as possible
to accurately represent how the probability to follow the expected rate evolves.

To get the maximum redshift after which we stop following the expected rate, we
took the redshift for which we are at a 0.3 on the interpolated cumulative
distribution function. We chose that value to have a systematic analytical
result that matches what we expect from the histograms; its impact on the work
after is to be studied.

\begin{table}[htbp!]
    \centering
    \caption{Number of SNe Ia effectively used for each survey, totalling 689
    Sne Ia. TO CORRECT FOR NEW ZMAX}
    \label{tab:sample}
    \begin{tabular}{c c}\hline
        \rowcolor{gray!15} Survey & Number of SNe Ia \\\hline
        SNf & 141 \\
        PS1 & 178 \\
        SDSS & 206 \\
        SNLS & 138 \\
        HST & 26 \\\hline
    \end{tabular}
\end{table}

@@

We used the LsSFR and stretch measurments from the SNf sample. The LsSFR has an
evolution with the redshift that is analytically known: calling $\de\z$ the
fraction of young stars and $\p\z$ the fraction of old ones, Rigault 2018 et +
find
\begin{equation}
    \frac{\de\z}{\p\z} \equiv \mathrm{LsSFR}\z = K\times \LF{1+z}^\f
\end{equation}
with $\f = 2.8$, and knowing $\de\z + \p\z = 1$:
\begin{align}
    \de\z & = \LF{K^{-1}\times \LF{1+z}^{-\f} +1}^{-1}\\
    \p\z & = \LF{K\times \LF{1+z}^{+\f} + 1}\um
\end{align}

\begin{figure*}[htbp!]
    \centering
    \subfigure[Raw data]{\label{fig:snf_data}
        \includegraphics[width=8.5cm]{Article_figures/BiGaussian_onlydata.pdf}}
    \subfigure[First implemented model]{\label{fig:3G2M2SSNF}
        \includegraphics[width=8.5cm]{Article_figures/BiGaussian.pdf}}
    \caption{Stretch des supernovae étudiées par la collaboration SNF en
    fonction de log(LsSFR). La couleur représente la probabilité pour une
supernova d'être issue d'un jeune progéniteur.}
    \label{fig:mod_first}
\end{figure*}

@@

        \rowcolor{gray!15} Modèle & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ &
        $\s_2$ \\\hline

@@

    \rowcolor{gray!15} Modèle & $\m$ & $\s$ \\\hline

@@

    \rowcolor{gray!15} Modèle & $\m$ & $\s_-$ & $\s_+$ \\\hline

@@

    \rowcolor{gray!15} Modèle & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ & $\s_2$ &
    $m_3$ & $\s_3$ \\\hline

@@

        \rowcolor{gray!15} Name & Description & Free param &
        $\ln\mathcal{L}$ & $AICc$ & $\D AICc$ & Proba \\\hline

>>>>

\documentclass[bibnumber]{aa}

\usepackage{hyperref}

\usepackage{color}
\newcommand{\mr}[1]{{\textcolor[rgb]{0.60,0.10,0.6}{#1}}}
\newcommand{\mri}[1]{{\textcolor{red}{#1}}}

@@

\author{N. Nicolas\inst{1}
    \and M. Rigault\inst{2}
    \and R. Graziani\inst{2}     
    \and M. Briday\inst{1}
    \and Y. Copin\inst{1}    
    \and Y. Kim\inst{1}
}

@@

%\begin{figure*}
%    \centering
%    \subfigure[Method]{\label{fig:zdet}
%        \includegraphics[width=8.5cm]{Article_figures/zmax_maglim_all.pdf}}
%    \subfigure[Result]{\label{fig:surveys_cuts}
%    \includegraphics[width=8.5cm]{A    \subfigure[Method]{\label{fig:zdet}
%rticle_figures/surveys_cuts_x.pdf}}
%    \subfigure[Method]{\label{fig:zdet}

%    \caption{(a) $z\St{max}$ determination from the magnitude limited data; (b) SDSS, %SNLS and PS1 samples cut at $z\St{max}$ as explained in
%    section \ref{sec:sam}. The data we use is in plain markers.}
%\end{figure*}

Type Ia supernovae (SNe Ia) are powerful cosmological distance indicators that
have enabled the discovery of the acceleration of the Universe's expansion
\citep{riess1998, permutter1999}. They remain today a key comological probe to
understand the properties of dark energy (DE) as it is the only tool able to
precisely map the recent expansion rate $z<0.5$, when DE is driving it
\citep[e.g.][]{scolnicastro2020}. They also are key to directly measure the
Hubble Constant (H$_0$) provided one can calibrate their absolute magnitude
\citep{riess2016, freedman2019}. Interestingly, the value of H$_0$ derived when
the SNeIa are anchored on Cepheids \citep[the SH0ES
project][]{riess2009,riess2016} is $\sim5~\sigma$ higher than what is predicted
by cosmic microwave background (CMB) data measured by Planck assuming the
standard $\L$CDM or when the SN luminosity is anchored at intermediate redshift
by the baryon accoustic oscillation (BAO) scale \citep{riess2019, ried2019,
planck, feeney}. While using the tip of the red giant branch technique in place
of the Cepheids seem to favor lower values of H$_0$ \citep{freedman,
recalibrationpaper} time delay measurements from strong lensing seem to also
favor high H$_0$ values \citep{wong2019}.

The H$_0$ tension has received a lot of attention as it could be a sign of new
fundamental physics. Yet, no simple solution is able to accomodate this H$_0$
tension when accounting for all other probes and the current most promising
scenario appears to be a burst of expansion at the matter-radiation decoupling
moment caused by a (fine tuned) early dark energy \citep{poulin}.

Alternatively, systematic effects affecting one or several of the aforementioned
analysis could also explain the tension. In \cite{rigault2015} we suggested that
SNe Ia from the cepheid calibrator sample differ significantly to the one from
the Hubble flow sample that enables to derive H$_0$. Indeed, Cepheid are very
young stars and the existence of Cepheids in a galaxy implies that the host is
star forming. In \citep{Rigault2015} we claimed that a more than 90\% of the
calibrator SNe Ia arise from a young-progenitor propulation while this
population only accounts for about half of the Hubble flow sample. Following
\cite{sullivan2010} and \citep{rigault2013}, we further showed that young and
old progenitor SNe Ia have a different magnitude of about 0.15 mag. This has
been confirmed for modern standardization technique
\citep[SALT2.4][]{guy2010,betoule2014} with high significance ($\sim 6\sigma$)
in \citep{rigault2018} as well as independently by \cite{roman2019} at $\sim
7\sigma$. As detailed in \citep{Rigault2015}, if the magnitude difference
between young and old SNe Ia is indeed of $0.15$ mag and if SNe Ia from the
calibrator sample is fully dominated by young SNeIa, the effect on H$_0$ of the
existance of these two SNe Ia populations is of 3.5\%. Accounting for the fact
that another environmental effect (the mass-step) is already included in the
SH0ES analysis, the net bias in the SH0ES analysis is expected closer to 3\%
\citep[see table 6 of ][]{rigault2015}.

SHOULD BE SHORTENED A LOT: However, the importance of this astrophysical bias in
the direct measurement of H$_0$ is still highly debated. We also highlight that
it could not explain the high value of H$_0$ derived by strong lensing. First,
\cite{jones2015} did not find any environmental bias when extending the
\citep{rigault2015} study. It was more recently reported in \citep{jones2019}
that indeed local host analysis has significant influence on the SN magnitude
but much more importantly than claimed in \citep{rigault2018} and with a
connection with the progenitor physics and/or a line of sight effect that is
unclear. More importantly, \cite{riess2016} attended to mimic the cepheid sample
selection function on the Hubble flow sample by removing SNe Ia from the latter
if they classificed their host as early-type or non-star forming. They measured
that doing so has no effect on the measurement of H$_0$.

% SERIES OF PAPER
This paper is part of a series of papers where we test the validity of the two
SNe Ia population models developped in \citep{rigault2018}.Briday et al. in prep
will present how using different environmental tracer is affecting our ability
to distinguish the two populations, and Rigault et al. in prep will further
study the connection with H$_0$. Here we analyze predictions concerning the
SNe Ia redshift drifts: (1) that the distribution of the SNIa stretch, a purely
intrinsic SNIa property, is evolving as a function of redshift and (2) that this
population-drift follows the relation that \citep{rigault2018} derived assuming
that one population is “young“ with a rate proportional to the star formation
rate in the Universe and the second is "old" with a rate proportional to the
stellar mass.

The concept of the SNeIa age dichotomy arised when the SNIa rate has first been
studied. \cite{Mannucci et al. 2005; Scannapieco & Bildsten 2005; Sullivan et
al. 2006; Aubourg et al. 2008} have shown that the relative SNe Ia rate in
galaxies could only be explained if two populations existed, one young,
following the host star formation activity, and one old following the host
stellar mass (the so called “prompt and delayed“ or “A+B“ model). In
\cite{rigault2018} we used the specific star formation rate at the SN location
(Local sSFR or LsSFR) to classify which are the prompts (those with a high
LsSFR) and which are the delayed (those with low LsSFR). Since the first SNe Ia
host analysis, the SN strech has been known to be strongly correlated with the
SN host properties \citep{Hamuy1996, Hamuy2000} and it has been extensively
confirmed since \citep[e.g.][]{Neill2009,Sullivan2010; Lampeitl2010; Kelly2010;
Gupta2011,D’Andrea2011; Childress2013,Rigault2013,Pan2014}. Following the “A+B“
model and the connection between SN stretch and host properties,
\citep{Howell20XX} first discussed the potential redshift drift of the SN strech
distribution. In this paper we revisit this analysis with the most recent SNe Ia
dataset and we test the validity of the two SNIa age populations model
developped by \citep{rigault2018}.

We present in section~\ref{sec:data} the sample we are using for this analysis,
which is based on the pantheon dataset \citep{scolnic2018a}. We discuss the
importance of obtaining a "complete" sample, i.e. reprensentative of the true
underlying SNe Ia distribution, and how we build one from the Pantheon sample.
We then present in section~\ref{sec:model} our modeling of the distribution of
stretch as a function of redshift based on the relative stretch distrubtion of
young and old SNeIa. Our results are presented in section~\ref{sec:result} where
we test if the distribution of stretch does evolve as a function of redshift and
if the age-model is in good agreement with this evolution. We discuss our
results and conclude in section~\ref{sec:conclusion}.

Throughout the paper we use the Planck 2015 cosmology from the astropy library.

@@

We base our analysis on the latest SNe Ia compilation: the Pantheon catalog from
\cite{scolnic2018a}. Using this dataset, we aim at probing the evolution of the
distribution of SNe Ia stretch in nature as a function of redshift. Because the
observed magnitude of a SN Ia is correlated with its stretch and its color, in a
magnitude limited survey such as those consituing most of the Pantheon sample,
the distribution of these lightcurve parameters is not representative of the
true underlying distribution above a certain limiting redshift. The first SNe Ia
that such a survey will miss because they will be too faint are indeed the
reddest and the lowest-stretch SNe Ia. If such a selection effect is not
accounted for, we might confuse true population drift we selection function or
conversally.

The figure~\ref{fig:maglim} shows the distribution of stretch for the magnitude
limited samples of the Pantheon data, i.e. PanStarrs \citep[PS1][]{ref}, the
Sloan Digital Sky Survey \citep[SDSS][]{ref} and the SuperNovae Legacy Survey
\citep[SNLS][]{astier2006}. An ellipse in the salt2.4 $(x_1, c)$ plan with major
axis $x_1=3$ and manor axis $c=0.3$ incapsulate the full distribution \citep[see
also][for a similar contour but using a less conservative
$c=0.2$]{campbell2013}. Assuming the average SN absolute magnitude ($x_1=0$,
$c=0$) is $M=-19.36$ \citep{kessler2009,scolnic2014} we can derive the absolute magnitude of
SNeIa along this ellipse given the standardisation coefficient from
\citep{scolnic2018a}: $\alpha=0.16$ and $\beta=3.14$.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Article_figures/zmax_maglim_all.pdf}
    \caption{SALT2.4 stretch and color lightcurve parameters of SNeIa from the
        SDSS, PS1 and SNLS samples from the pantheon catalog. The individual SN
        data are shown as blue dots and a 2D histogram is shown in grey to
        highlight point density. The ellipse encapsulating all the SN data is
        displayed colored by the effective standardized absolute magnitude using
    the $\alpha$ and $\beta$ standardisation coefficents from
\citep{scolnic2018}.}
    \label{fig:maglim}
\end{figure}

The faintest SNIa is that with $x_1=-1.66$ and $c=0.25$ and has an absolute \mr{standardized} magnitude at peak \mr{in Bessel-b band} of $-18.31$ mag, which leads to limiting asbolute magnitude of
$-18.00$ if we want to be able to observe it typically a week before and 10 days after peak.  
Therefore, given the $5\sigma$ point source detection magnitude
limit of a magnitude limited survey, one can derive the maximum redshift above which the faintest SNeIa will start to be missed.

\mr{SNLS typically acquires SNeIa in the redshift range $0.4<z<0.8$. At these redshifts the rest-frame Bessel-b band roughly corresponds to the SNLS-$i$ filter, that has a 24.8 mag $5\sigma$ depth \footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT final release website.}}. This converts to a $z_{lim}=0.60$, in perfect agreement with \citep{neill2006} and \citep{perrett2010}.
Similarly, PS1 observing SNeIa in the range $0.2<z<0.4$, their $g$-band $5\sigma$ depth is 23.1 mag \citep{rest2014}, corresponding $z_{lim}=0.30$.
The fig.~6 of \citep{scolnic2018} suggests a more conservative $z_{lim}$ of 0.27 for the PS1 catalog. In similar redshift range, SDSS has a limiting magnitude of 22.5 \citep{dilday2008,sake2008,sako2014}, which would lead to a $z_max=0.24$. However, the SDSS survey were more sensitive to limited spectroscopic ressources. \cite{kessler2009} presents that during year-1 of SDSS, SNeIa with $r-mag<20.5$ where favored for spectroscopic follow up, corresponding to the redshift cut at $0.15$. For the rest of the SDSS survey, additional spectroscopic ressources where used, such that \cite{kessler2009} and  \cite{dilday2008} of show a relative completness up to $z_{lim}=0.2$. For the rest of the analysis, we will use $z_{lim}=0.2$ as baseline magnitude.}

\mr{In Section~\ref{sec:testzmax}, we test the impact on using more conservative redshift limits ($z_{lim}=0.15$ for SDSS, $z_{lim}=0.27$ for PS and $z_{lim}=0.55$ for SNLS, following Fig.~14 of \citealt{perrett2010}). 
}

In addition, we use the SNe Ia sample of from the nearby supernovae factory
\citep[SNfactory][]{aldering2004} published in \citep{rigault2018}. As the SNe
search were much deeper than the spectrophotometric follow-up, SNfactory SNe Ia
within a redshift range of $0.02<z<0.09$ should also be a random sampling of the
underlying SN population. The HST sample from
Pantheon follows the same logic and we therefore kept it entirely \citep{FIND BACK THE REF}.

Our "complete sample" is therefore made of 656 SNeIa whose composition is
detailed in table~\ref{tab:sample}.

\begin{table}
    \centering
    \caption{Origin of the SNeIa in our sample and redshift limit (if
    applicable) of each surveys.}
    \label{tab:sample}
    \begin{tabular}{l c c}
    \hline\hline\\[-0.8em]
        Survey & N$_{\mathrm{SN}}$  & z$_{lim}$ \\[0.15em]
        \hline\\[-0.8em]
        SNf & 141 & --\\[0.30em]
        SDSS & 167 & 0.20\\[0.30em]
        PS1 & 160 & 0.30 \\[0.30em]        
        SNLS & 102 & 0.60\\[0.30em]
        HST & 26 & --\\[0.30em]
        \hline
    \end{tabular}
\end{table}

%\mri{COMPARE HERE WITH LITERATURE. E.G. FIG. 6 OF SCOLNIC 2018, PS zmax AROUND0.27}

%\mri{Perrett et al. 2010 | SNLS | zmax ="z $\sim$ 0.6"}

%\mri{Dilday et al. 2008 | SDSS | sure good at z<0.12, fig. 10 suggest still okat z=0.2}

%\mri{Kessler 2009 "For the SDSS-II, the image-subtraction
%pipeline efficiency (	subtr) is complete up to a redshift of z ∼ 0.2"}

%\mri{Kesslet 2009 | "For the SDSS-II and SNLS samples, the cutoff redshifts are
%estimated to be 0.15 and 0.65, respectively."}
% ADDIND: PS cut from pantheon close to 0.27 see Fig. 6 of Scolnic 2018

\mri{CHECK LITERATURE CAN WE USE THE CfA SAMPLES ???}

@@

We used the LsSFR and stretch measurments from the SNf sample. The LsSFR has an
evolution with the redshift that is analytically known: calling $\de\z$ the
fraction of young stars and $\p\z$ the fraction of old ones, Rigault 2018 et +
find
\begin{equation}
    \frac{\delta\z}{\psi\z} \equiv \mathrm{LsSFR}\z = K\times \LF{1+z}^\varphi
\end{equation}
with $\f = 2.8$, and knowing $\de\z + \p\z = 1$:
\begin{align}
    \delta\z & = \left{K^{-1}\times \LF{1+z}^{-\f} +1\right}^{-1}\\
    \psi\z & = \LF{K\times \LF{1+z}^{+\f} + 1}\um
\end{align}

%\begin{figure*}
%    \centering
%    \subfigure[Raw data]{\label{fig:snf_data}
%        \includegraphics[width=8.5cm]{Article_figures/BiGaussian_onlydata.pdf}}
%    \subfigure[First implemented model]{\label{fig:3G2M2SSNF}
%        \includegraphics[width=8.5cm]{Article_figures/BiGaussian.pdf}}
%    \caption{Stretch des supernovae étudiées par la collaboration SNF en
%    fonction de log(LsSFR). La couleur représente la probabilité pour une
%supernova d'être issue d'un jeune progéniteur.}
%    \label{fig:mod_first}
%\end{figure*}

@@

        Model & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ &
        $\s_2$ \\\hline

@@

    Model & $\m$ & $\s$ \\\hline

@@

    Model & $\m$ & $\s_-$ & $\s_+$ \\\hline

@@

    Model & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ & $\s_2$ &
    $m_3$ & $\s_3$ \\\hline

@@

        Name & Description & Free param &
        $\ln\mathcal{L}$ & $AICc$ & $\D AICc$ & Proba \\\hline
